Precise Asymptotics of Transferring from Different Regression in High Dimensions

### 2021-06-14 revision_journal/intro.tex
restructure
sec 2/3 wip

\begin{figure*}[!t]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.45\textwidth,valign=t]{figures/mtl_model_arch.pdf}
		\caption{A hard parameter sharing architecture}
		\label{fig_intro_arch}
	\end{subfigure}\hfill
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.745\textwidth,valign=t]{figures/sample_ratio_c2_400.eps}
		\caption{Varying sample size ratio}
		\label{fig_intro_sample_size_b}
	\end{subfigure}
	\vspace{-0.1in}
	\caption{
	Left: an illustrative picture of HPS.
	Right: an illustrative example of using HPS for two tasks $X_1, Y_1$ and $X_2, Y_2$ with sample size $n_1, n_2$, respectively.
	Increasing $n_1/n_2$ decreases task two's prediction loss initially but increase afterward. This phenomenon occurs due to different bias-variance tradeoffs as $n_1/n_2$ increases. Our result provides an estimated loss (solid line) that accurately matches the empirical loss (dotted line).
	See Section \ref{sec_simulation} for the precise setting.}
	\label{fig_intro_sample_size}
\end{figure*}

@@

\subsection{Setup and definitions}
Consider the transfer learning setting with two data sets. We denote their sample sizes by $n_1$ and $n_2$. For $i=1,2,$ Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates. We assume that the label vector $Y^{(i)} \in \real^{n_i}$ for $X^{(i)}$ follows a linear model plus random noise:
$$Y^{(i)}= X^{(i)}\beta^{(i)} + \varepsilon^{(i)},\quad i=1,2.$$
In this paper, we make the following assumptions on feature covariates $X^{(i)}$ and the noise vectors $\varepsilon^{(i)}$. We remark that all these assumptions are natural in high-dimensional statistics.

We assume that the row vectors of $X^{(i)}$ are i.i.d. centered random vectors with  population covariance matrix $\Sigma^{(i)}$. More precisely, let
\be\label{XofZ}X^{(i)} = Z^{(i)} (\Sigma^{(i)})^{1/2} \in \real^{n_i\times p},\quad i=1,2, \ee
where each $\Sigma^{(1)}$ and $\Sigma^{(2)}$ are $p\times p$ deterministic positive definite symmetric matrices, and $Z^{(1)}=(z^{(1)}_{ij})$ and $Z^{(2)}=(z^{(2)}_{ij})$ are $n_1\times p$ and $n_2\times p$ random matrices with real i.i.d. entries satisfying
\begin{equation}\label{assm1}
\mathbb{E} z^{(1)}_{ij}=\mathbb{E} z^{(2)}_{ij} =0, \ \quad \ \mathbb{E} \vert z^{(1)}_{ij} \vert^2=\mathbb{E} \vert z^{(2)}_{ij} \vert^2 =1,
\end{equation}

Another commonly used transfer learning estimator is defined as the minimizer of the total loss function \cite{bastani2020predicting, li2020transfer}:
$$\hat \beta^{\rm{TL}} := \argmin\left[ \norm{X^{(1)} \beta - Y^{(1)}}^2 + \norm{X^{(2)} \beta - Y^{(2)}}^2\right].$$

Given a solution from minimizing $f(A, B)$, denoted by $(\hat{A}, \hat{B})$  (which we will specify below), let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the HPS estimator for task $i$.
The critical questions are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

\subsection{Main results}
Our first result (Theorem \ref{thm_many_tasks}) applies to multi-label prediction settings where all datasets have the same features (and sample size), and we want to make several predictions for every input (cf. examples in \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp bias-variance decomposition of its (out-of-sample) prediction loss for any task.
This setting is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
%We show that the prediction loss of HPS admits a clean bias-variance decomposition.
Our result implies that when all tasks have the same features but different labels, for any task, HPS helps reduce the task's variance compared to single-task learning but increases bias.

@@

\medskip
\noindent\textbf{Proof techniques.}
There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimizer of $f(A, B)$ and its population version (cf. equation \eqref{eq_A_star}).
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. We show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}).
We use recently developed techniques from random matrix theory to show a sharp convergence rate.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}.
%\HZ{add several sentences on the technical insight}
One limitation of our analysis is that in Example \ref{ex_sample_ratio}, there is an error term that can result in vacuous bounds for very small $n_1$ (cf. equation \eqref{cor_MTL_error}).
We believe our result has provided significant initial insights, and it is an interesting question to tighten our result.
See Section \ref{sec_conclude} for more discussions of the technical challenge.

@@

\subsection{Organizations}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation}, we validate our results in simulations.
In Section \ref{sec_conclude}, we summarize our work and discuss future work.
Section \ref{sec_text} describes our study on text classification tasks.
Section \ref{app_proof_error_same_cov}, \ref{appendix RMT}, and \ref{app_iso_cov} present proofs of our results.

>>>>

\begin{figure*}[!t]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.745\textwidth,valign=t]{figures/sample_ratio_c2_400.eps}
		\caption{Covariate shift}
		\label{fig_intro_arch}
	\end{subfigure}\hfill
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.745\textwidth,valign=t]{figures/sample_ratio_c2_400.eps}
		\caption{Varying sample size ratio}
		\label{fig_intro_sample_size_b}
	\end{subfigure}
	\vspace{-0.1in}
	\caption{
	Left: an illustrative picture of HPS.
	Right: an illustrative example of using HPS for two tasks $X_1, Y_1$ and $X_2, Y_2$ with sample size $n_1, n_2$, respectively.
	Increasing $n_1/n_2$ decreases task two's prediction loss initially but increase afterward. This phenomenon occurs due to different bias-variance tradeoffs as $n_1/n_2$ increases. Our result provides an estimated loss (solid line) that accurately matches the empirical loss (dotted line).
	See Section \ref{sec_simulation} for the precise setting.}
	\label{fig_intro_sample_size}
\end{figure*}

@@

\subsection{Summary of Results}

Our first result (Theorem \ref{thm_many_tasks}) applies to multi-label prediction settings where all datasets have the same features (and sample size), and we want to make several predictions for every input (cf. examples in \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp bias-variance decomposition of its (out-of-sample) prediction loss for any task.
This setting is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
%We show that the prediction loss of HPS admits a clean bias-variance decomposition.
Our result implies that when all tasks have the same features but different labels, for any task, HPS helps reduce the task's variance compared to single-task learning but increases bias.

@@

%\medskip
%\noindent\textbf{Proof techniques.}
%There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
%We carefully keep track of the concentration error between the global minimizer of $f(A, B)$ and its population version (cf. equation \eqref{eq_A_star}).
%The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. We show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}).
%We use recently developed techniques from random matrix theory to show a sharp convergence rate.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}.
%\HZ{add several sentences on the technical insight}
%One limitation of our analysis is that in Example \ref{ex_sample_ratio}, there is an error term that can result in vacuous bounds for very small $n_1$ (cf. equation \eqref{cor_MTL_error}).
%We believe our result has provided significant initial insights, and it is an interesting question to tighten our result.
%See Section \ref{sec_conclude} for more discussions of the technical challenge.

@@

\paragraph{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation}, we validate our results in simulations.
In Section \ref{sec_conclude}, we summarize our work and discuss future work.
Section \ref{sec_text} describes our study on text classification tasks.
Section \ref{app_proof_error_same_cov}, \ref{appendix RMT}, and \ref{app_iso_cov} present proofs of our results.

### 2021-06-14 revision_journal/results.tex
sec 2/3 wip

\subsection{Assumptions}
Furthermore, we assume that the entries $z^{(1)}_{ij}$ and $z^{(2)}_{ij}$ have finite $\varphi$-th moment for some constant $\varphi>4$:
\be \label{conditionA2}
\mathbb{E} \vert z^{(1)}_{ij} \vert^\varphi  \le \tau^{-1},\quad \mathbb{E} \vert  z^{(2)}_{ij} \vert^\varphi   \le \tau^{-1}
\ee
for a small constant $\tau>0$. We assume that $\Sig^{(i)}$ has eigendecomposition
\be\label{eigen}
\Sig^{(i)}= O_i\Lambda_i O_i^\top, \quad  \Lambda_1=\text{diag}(\si_1^{(i)}, \ldots, \si^{(i)}_n),  \quad i=1,2,
\ee
where $O_i$ is the eigenmatrix and the eigenvalues satisfy that
\begin{equation}\label{assm3}
 \tau \le  \si^{(i)}_p \le\cdots\le \si^{(i)}_2 \le \si^{(i)}_1 \le \tau^{-1} \quad i=1,2.
\ee
%for a small constant $\tau>0$.

@@

%\subsection{Transfer learning estimators}

@@

On the other hand, much less is known about the predication loss of the HPS estimator. In particular, its exact dependence on the model parameters, including the model bias, noise variance, sample sizes, and population covariance matrices, is not well-understood so far. %\FY{What is known so far about this topic?}
In Section \ref{sec_HPS_loss}, we give a rigorous analysis of the bias term \eqref{Lbias} and variance term \eqref{Lvar} in two different settings: (i) independent $X^{(1)}$ and $X^{(2)}$ with different sample sizes and same population covariance matrices $\Sigma^{(1)}=\Sigma^{(2)}$; (ii) independent $X^{(1)}$ and $X^{(2)}$ with different sample sizes and different population covariance matrices. In case (i), we provide the exact asymptotics for both the bias and variance terms, together with almost sharp convergence rates. We will use these results to illustrate the effects of the \emph{bias-variance tradeoff} and \emph{varying sample sizes}. In case (ii), we will prove an exact asymptotic variance limit, and employ the result to show the intricate effect of the covariate shift. On the other hand, in case (ii), we will provide an estimate on the bias term, which becomes exact only when $n_1\gg p$. %Here case (ii) is actually more general than (i)
%In particular, we will use our results to analyze the effect of the following factors: \emph{bias-variance tradeoff}, \emph{sample sizes}, and \emph{covariate shift}.
In Section \ref{sec_same}, we will consider a simple extension to a multi-task learning setting with more than two tasks and  the same covariates for all tasks. We will give the asymptotic limit of the predication loss for the HPS estimator, which provides useful insights into the more rich multi-task learning theory. Due to length constraint, we will explore the multi-task learning setting in greater details in other works.

@@

\paragraph{Examples.}

@@

\section{Asymptotics under Imbalanced Sample Sizes} \label{sec_sizeratio}

@@

 %{How does hard parameter sharing scale with sample sizes and covariate shift $M$?} One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift. Next, we illustrate how varying them impact the prediction loss.
\begin{example}%[Sample sizes]
\label{ex_sample_ratio}
	%We first consider the impact of varying sample sizes.
	%Consider the random-effect model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.
	%We again consider the random effect model in Example \ref{ex_same_cov} with $t=2$.
	Suppose each $\beta^{(i)}$ consists of two random components, one that is shared among the two tasks and one that is task-specific. More precisely, we let
$$\beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2, $$
where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
and $\wt \beta^{(i)}$ denotes the $i$-th
%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} d^2$. First, we show that if $\|\beta_0\|^2 \gg d^2$, then the global minimizer $\hat a$ of the function $g(a)$ in \eqref{eq_mtl_A12} is close to $1$.
	The proof of Proposition \ref{lem_hat_v} will be given in Appendix \ref{app_iso_cov_prop}.

>>>>

\subsection{Models and Assumptions}

Consider the transfer learning setting with two data sets. We denote their sample sizes by $n_1$ and $n_2$. For $i=1,2,$ Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates. We assume that the label vector $Y^{(i)} \in \real^{n_i}$ for $X^{(i)}$ follows a linear model plus random noise:
$$Y^{(i)}= X^{(i)}\beta^{(i)} + \varepsilon^{(i)},\quad i=1,2.$$
In this paper, we make the following assumptions on feature covariates $X^{(i)}$ and the noise vectors $\varepsilon^{(i)}$. We remark that all these assumptions are natural in high-dimensional statistics.

We assume that the row vectors of $X^{(i)}$ are i.i.d. centered random vectors with  population covariance matrix $\Sigma^{(i)}$. More precisely, let
\be\label{XofZ}X^{(i)} = Z^{(i)} (\Sigma^{(i)})^{1/2} \in \real^{n_i\times p},\quad i=1,2, \ee
where each $\Sigma^{(1)}$ and $\Sigma^{(2)}$ are $p\times p$ deterministic positive definite symmetric matrices, and $Z^{(1)}=(z^{(1)}_{ij})$ and $Z^{(2)}=(z^{(2)}_{ij})$ are $n_1\times p$ and $n_2\times p$ random matrices with real i.i.d. entries satisfying
\begin{equation}\label{assm1}
\mathbb{E} z^{(1)}_{ij}=\mathbb{E} z^{(2)}_{ij} =0, \ \quad \ \mathbb{E} \vert z^{(1)}_{ij} \vert^2=\mathbb{E} \vert z^{(2)}_{ij} \vert^2 =1,
\end{equation}

Another commonly used transfer learning estimator is defined as the minimizer of the total loss function \cite{bastani2020predicting, li2020transfer}:
$$\hat \beta^{\rm{TL}} := \argmin\left[ \norm{X^{(1)} \beta - Y^{(1)}}^2 + \norm{X^{(2)} \beta - Y^{(2)}}^2\right].$$

Given a solution from minimizing $f(A, B)$, denoted by $(\hat{A}, \hat{B})$  (which we will specify below), let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the HPS estimator for task $i$.
The critical questions are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

Furthermore, we assume that the entries $z^{(1)}_{ij}$ and $z^{(2)}_{ij}$ have finite $\varphi$-th moment for some constant $\varphi>4$:
\be \label{conditionA2}
\mathbb{E} \vert z^{(1)}_{ij} \vert^\varphi  \le \tau^{-1},\quad \mathbb{E} \vert  z^{(2)}_{ij} \vert^\varphi   \le \tau^{-1}
\ee
for a small constant $\tau>0$. We assume that $\Sig^{(i)}$ has eigendecomposition
\be\label{eigen}
\Sig^{(i)}= O_i\Lambda_i O_i^\top, \quad  \Lambda_1=\text{diag}(\si_1^{(i)}, \ldots, \si^{(i)}_n),  \quad i=1,2,
\ee
where $O_i$ is the eigenmatrix and the eigenvalues satisfy that
\begin{equation}\label{assm3}
 \tau \le  \si^{(i)}_p \le\cdots\le \si^{(i)}_2 \le \si^{(i)}_1 \le \tau^{-1} \quad i=1,2.
\ee
%for a small constant $\tau>0$.

@@

\subsection{Transfer Learning Estimators}

@@

On the other hand, much less is known about the predication loss of the HPS estimator. In particular, its exact dependence on the model parameters, including the model bias, noise variance, sample sizes, and population covariance matrices, is not well-understood so far. %\FY{What is known so far about this topic?}
In Section \ref{sec_HPS_loss}, we give a rigorous analysis of the bias term \eqref{Lbias} and variance term \eqref{Lvar} in two different settings: (i) independent $X^{(1)}$ and $X^{(2)}$ with different sample sizes and same population covariance matrices $\Sigma^{(1)}=\Sigma^{(2)}$; (ii) independent $X^{(1)}$ and $X^{(2)}$ with different sample sizes and different population covariance matrices. In case (i), we provide the exact asymptotics for both the bias and variance terms, together with almost sharp convergence rates. We will use these results to illustrate the effects of the \emph{bias-variance tradeoff} and \emph{varying sample sizes}. In case (ii), we will prove an exact asymptotic variance limit, and employ the result to show the intricate effect of the covariate shift. On the other hand, in case (ii), we will provide an estimate on the bias term, which becomes exact only when $n_1\gg p$. %Here case (ii) is actually more general than (i)
%In particular, we will use our results to analyze the effect of the following factors: \emph{bias-variance tradeoff}, \emph{sample sizes}, and \emph{covariate shift}.
In Section \ref{sec_same}, we will consider a simple extension to a multi-task learning setting with more than two tasks and  the same covariates for all tasks. We will give the asymptotic limit of the predication loss for the HPS estimator, which provides useful insights into the more rich multi-task learning theory. %Due to length constraint, we will explore the multi-task learning setting in greater details in other works.

@@

 %{How does hard parameter sharing scale with sample sizes and covariate shift $M$?} One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift. Next, we illustrate how varying them impact the prediction loss.
\begin{example}%[Sample sizes]
\label{ex_sample_ratio}
	%We first consider the impact of varying sample sizes.
	%Consider the random-effect model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.
	%We again consider the random effect model in Example \ref{ex_same_cov} with $t=2$.
	Suppose each $\beta^{(i)}$ consists of two random components, one that is shared among the two tasks and one that is task-specific. More precisely, we let
$$\beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2, $$
where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
and $\wt \beta^{(i)}$ denotes the $i$-th
%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} d^2$.

@@

\section{Asymptotics under Sample Size Imbalance}\label{sec_sizeratio}

@@

\paragraph{Examples.}
First, we show that if $\|\beta_0\|^2 \gg d^2$, then the global minimizer $\hat a$ of the function $g(a)$ in \eqref{eq_mtl_A12} is close to $1$.
	The proof of Proposition \ref{lem_hat_v} will be given in Appendix \ref{app_iso_cov_prop}.
